CrypTool Script \\ Mathematics and Cryptography

### 2003-03-27 script/en/chapters/cryptomethods.tex
more be updates

% --------------------------------------------------------------------------
\subsection{Encryption}

@@

One encryption procedure has been proved to be secure -- the {\em One
Time Pad}\index{One ­Time ­Pad}. However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be just as long
as the message to be protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.\par \vskip + 3pt

@@

We basically distinguish between symmetric and asymmetric encryption procedures.

% --------------------------------------------------------------------------
\subsubsection[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} v1.3 you can execute the following modern
    symmetric encryption algorithms 
    (using the menu {\bf Crypt \textbackslash{} Symmetric}): \\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm).
  }

@@

@@

The most well-known symmetric encryption procedure is the \index{DES} DES-algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing all possible keys until the right one is found ({\em brute-force-attack}).
\index{Attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.\par \vskip + 3pt

Up-to-the-minute procedure is the symmetric AES standard. The associated
Rijndael algorithm was declared winner of the AES award on 2 October 2000
and thus succeeds the DES procedure.

More details about the AES algorithms can be found within the 
online help of CrypTool\index{CrypTool}%
\footnote{%
      CrypTool Online help\index{CrypTool}: the index head-word {\bf AES}
      leads to the 3 help pages: {\bf AES candidates}, 
      {\bf The AES winner Rijndael} and 
      {\bf The Rijndael encryption algorithm}.
  }
.

@@

Cryptanalysts already label methods as ``academic success'' or as ``cryptanalytic attack'' if they are theoretically faster than the complete testing of all keys (brute force analysis). In the case of AES with the maximal key length (256 bit) exhaustive key search on average needs $2^{255}$ encryption operations. A cryptanalytic attack needs to be better than this. At present between $2^{75}$ and $2^{90}$ encryption operations are estimated to be performable for organizations, for example a security agency.

@@

The discussion is very controversial: Don Coppersmith (one of the inventors of DES) for example queries the practicability of the attack because XLS would provide no solution for AES \cite{Coppersmith2002}. This concludes that then the optimization of Murphy and Robshaw \cite{Robshaw2002b} would not work.

% --------------------------------------------------------------------------
\subsubsection[Asymmetric encryption]
{Asymmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} v1.3 you can execute RSA encryption
    and decryption (using the menu {\bf Crypt \textbackslash{} Asymmetric}).
  }

@@

If Alice wants to communicate with Bob, then she finds Bob's public key 
in the directory and uses it to encrypt her message to him. She then sends
this cipher text to Bob, who is then able to decrypt it again using his 
secret key. As only Bob knows his secret key, only he can decrypt 
messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

@@

The most well-known asymmetric procedure is the \index{RSA} RSA algorithm,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signatures!digital}Signature
Algorithm).

% --------------------------------------------------------------------------
% \newpage
\subsubsection[Hybrid procedures]
{Hybrid procedures\footnotemark}
\footnotetext{%
Within CrypTool\index{CrypTool} v1.3 you can get a visualization of this
technique using the menu {\bf Crypt \textbackslash{} Hybrid Demonstration}: 
this dialogue shows the single steps and its dependencies with concrete
numbers.
}\index{Hybrid procedures}

@@

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} v1.3 you can check out
different random number generators using the menu {\bf Indiv. Procedures 
\textbackslash{} Generate Random Numbers}. 
Using the menu{\bf Analysis \textbackslash{} Random Tests} you can apply
different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo number generators. Only the integrated Secude generator 
involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

% --------------------------------------------------------------------------
\subsubsection{Further details}

Beside the information you can find in many books and on a good number 
of websites the online help of CrypTool\index{CrypTool} also offers very
many details about the symmetric and asymmetric encryption methods.

@@

\begin{enumerate}
  \item[] \href{http://www.cryptosystem.net/aes}
               {\texttt{http://www.cryptosystem.net/aes}}
  \item[] \href{http://www.minrank.org/aes/}
               {\texttt{http://www.minrank.org/aes/}}
\end{enumerate}

>>>>

This chapter introduces the topic in a more descriptive way without using too much mathematics.

% --------------------------------------------------------------------------
% \subsection{Encryption}

@@

One encryption procedure has been proved to be secure -- the {\em One
Time Pad}\index{One ­Time ­Pad}. However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be at least as long
as the message being protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.\par \vskip + 3pt

@@

The book of Bruce Schneier \cite{Schneier1996} offers a very good overview about the different algorithms. We basically distinguish between symmetric and asymmetric encryption procedures.

% --------------------------------------------------------------------------
\subsection[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} v1.3 you can execute the following modern
    symmetric encryption algorithms 
    (using the menu {\bf Crypt \textbackslash{} Symmetric}): \\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm).
  }

@@

All classical methods are of this type. Examples can be found within 
CrypTool or in \cite{Nichols1996}. Now we want to consider more modern 
mechanisms.

@@

The most well-known symmetric encryption procedure is the \index{DES} DES-algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing (almost) all possible keys until the right one is found ({\em brute-force-attack}).
\index{Attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.\par \vskip + 3pt

Up-to-the-minute procedure is the symmetric AES standard. The associated
Rijndael algorithm was declared winner of the AES award on 2 October 2000
and thus succeedes the DES procedure.

More details about the AES algorithms can be found within the 
online help of CrypTool\index{CrypTool}%
\footnote{%
      CrypTool online help\index{CrypTool}: the index head-word {\bf AES}
      leads to the 3 help pages: {\bf AES candidates}, 
      {\bf The AES winner Rijndael} and 
      {\bf The Rijndael encryption algorithm}.
  }.

@@

Cryptanalysts already label methods as ``academic success'' or as ``cryptanalytic attack'' if they are theoretically faster than the complete testing of all keys (brute force analysis). In the case of AES with the maximal key length (256 bit) exhaustive key search on average needs $2^{255}$ encryption operations. A cryptanalytic attack needs to be better than this. At present between $2^{75}$ and $2^{90}$ encryption operations are estimated to be performable only just for organizations, for example a security agency.

@@

The discussion is very controversial at the moment: Don Coppersmith (one of the inventors of DES) for example queries the practicability of the attack because XLS would provide no solution for AES \cite{Coppersmith2002}. This concludes that then the optimization of Murphy and Robshaw \cite{Robshaw2002b} would not work.

% --------------------------------------------------------------------------
\subsubsection{Current status of brute-force attacks on 
symmetric algorithms (RC5)}
\index{brute-force attack}
\index{RC5}

The current status on brute-force attacks on symmetric encryption algorithms can be explained with the block cipher RC5.

Brute-force (exhaustive search, trial-and-error) means to completely examine all keys of the key space: so no special analysis methods have to be used. Instead, the cipher text is decrypted with all possible keys and for each resulting text it is checked, whether this is a meaningful clear text. A key length of 64 bit means at most 2^64 = 18,446,744,073,709,551,616 or about 18 trillion (GB) / 18 quintillion (US)  keys to check\index{CrypTool}%
\footnote{%
    With CrypTool\index{CrypTool} v1.3 you also can try brute-force attacks
    of modern symmetric algorithms 
    (using the menu {\bf Analysis \textbackslash{} Ciphertext-Only}): 
    to achieve a result in a appropriate time with a single PC you should 
    mark not more than 20 bit of the key as unknown.
}.

Companies like RSA Security provide so-called cipher challenges in order to quantify the security offered by well-known symmetric ciphers as DES, Triple-DES or RC5 (\href{http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}{\tt http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}). They offer prizes for those who manage to decipher cipher texts, encrypted with different algorithms and different key lengths and unveil the symmetric key (under controlled conditions). So theoretical estimates can be confirmed.

It is well-known, that the "old" standard algorithm DES with a fixed key length of 56 bit is no more secure: this was demonstrated already in January 1999 by the Electronic Frontier Foundation (EFF). With their specialized computer Deep Crack they cracked a DES encrypted message within less than a day (\href{http://www.rsasecurity.com/rsalabs/challenges/des3/index.html}{\tt http://www.rsasecurity.com/rsalabs/challenges/des3/index.html}).

The current record for strong symmetric algorithms unveiled a key 64 bit long. The algorithm used was RC5, a block cipher with variable key size.

The RC5-64 challenge has been solved by the distributed.net team after 5 years (\href{http://distributed.net/pressroom/news-20020926.html}{\tt http://distributed.net/pressroom/news-20020926.html}).  At all, 331,252 individuals co-operated over the internet to find the key. More than 15 trillion (GB) / 15 quintillion (US)  keys were checked, until they found the right key.

This makes clear, that symmetric algorithms (even if they have no cryptographical weakness) using keys of size 64 bit are no more appropriate to keep sensible data private.

Similar cipher challenges are there for asymmetric algorithms (please see chapter 3.11.4).

% --------------------------------------------------------------------------
\subsection[Asymmetric encryption]
{Asymmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} v1.3 you can execute RSA encryption
    and decryption (using the menu {\bf Crypt \textbackslash{} Asymmetric}).
  }

@@

If Alice\index{Alice}%
\footnote{%
      In order to describe cryptographic protocols the participants
      are often named Alice, Bob, \cdots (see \cite[S. 23]{Schneier1996}). 
      Alice and Bob perform all 
      2-person-protocols. Alice will initiate all protocols and 
      Bob answers. The attackers are named Eve (eavesdropper) and
      Mallory (malicious active attacker).
  } wants to communicate with Bob, then she finds Bob's public key 
in the directory and uses it to encrypt her message to him. She then sends
this cipher text to Bob, who is then able to decrypt it again using his 
secret key. As only Bob knows his secret key, only he can decrypt 
messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

@@

The most well-known asymmetric procedure is the \index{RSA} 
RSA algorithm\index{CrypTool}%
\footnote{%
The RSA algorithm is extensively described in chapter 3.10 and later
within this script. 
The RSA cryptosystem can be executed in many variations with 
CrypTool\index{CrypTool} v1.3 (using the menu 
{\bf Individual Procedures \textbackslash{} RSA Demonstration \textbackslash{}
RSA Cryptosystem}). 
The topical research results concerning RSA are described in chapter 3.11.
}%
,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signatures!digital}Signature
Algorithm).

% --------------------------------------------------------------------------
% \newpage
\subsection[Hybrid procedures]
{Hybrid procedures\footnotemark}
\footnotetext{%
Within CrypTool\index{CrypTool} v1.3 you can get a visualization of this
technique using the menu {\bf Crypt \textbackslash{} Hybrid Demonstration}: 
this dialogue shows the single steps and its dependencies with concrete
numbers.
}\index{Hybrid procedures}

@@

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} v1.3 you can check out
different random number generators using the menu {\bf Indiv. Procedures 
\textbackslash{} Generate Random Numbers}. 
Using the menu{\bf Analysis \textbackslash{} Random Tests} you can apply
different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude generator 
involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

% --------------------------------------------------------------------------
\subsection{Further details}

Beside the information you can find in the following chapters, many other
books and on a good number of websites the online help of 
CrypTool\index{CrypTool} also offers very many details about the 
symmetric and asymmetric encryption methods.

@@

\begin{enumerate}

\item AES or Rijndael Cryptosystem \\
        \href{http://www.cryptosystem.net/aes}
         {\tt http://www.cryptosystem.net/aes} \\
	\href{http://www.minrank.org/aes/}
         {\tt http://www.minrank.org/aes/}

  \item AES Discussion Groups at NIST \\
	\href{http://aes.nist.gov/aes}{\tt http://aes.nist.gov/aes}

  \item distributed.net:~\glqq RC5-64 has been solved'' \\
        \href{http://distributed.net/pressroom/news-20020926.html}
         {\tt http://distributed.net/pressroom/news-20020926.html}

  \item RSA: ~\glqq The RSA Secret Key Challenge'' \\
       \href{http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}
        {\tt http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}

  \item RSA: ~\glqq DES Challenge'' \\
        \href{http://www.rsasecurity.com/rsalabs/challenges/des3/index.html}
         {\tt http://www.rsasecurity.com/rsalabs/challenges/des3/index.html}

  \item Further Links can be found at the CrypTool Homepage \\
        \href{http://www.cryptool.org}
         {\tt http://www.cryptool.org}

\end{enumerate}

### 2003-03-27 script/en/chapters/ellipticcurves.tex
be updates
more be updates

\newpage
\section{Elliptic Curves}
\index{Elliptic curves}
\hypertarget{ellcurve}{}
(Filipovics B. / B"uger M. / Esslinger B. / Oyono R., April 2000, Updates: Dec. 2001, June 2002, Mar. 2003)

@@

There are factorisation algorithms based on elliptic curves
\footnote{In 1987 H.W. Lenstra published
a factorisation algorithm, based on elliptic curves (see \cite{Lenstra1987}).
The biggest compound number currently factorised with elliptic curves is 
the number $ 628^{59}-1, $ which has 55 decimal digits. It was
found Oct. 6th, 2001 by M. Izumi 
(See \hyperlink{Lenstra2}{ECMNET}\index{ECMNET}).
}. 
More precisely, these procedures exploit the fact that elliptic curves can
be defined over ${\mathbb Z}_n$ ($n$ composite number). Elliptic curves 
over ${\mathbb Z}_n$ do not form a group, because not every point on such 
an elliptic curve has an inverse point. This is connected with the fact 
that - if $n$ is a composite
number - there exist elements in ${\mathbb Z}_n$ that do not have an inverse
with respect to multiplication mod $n$. In order to add two points on an
elliptic curve over ${\mathbb Z}_n$, we can calculate in the same way as on
elliptic curves over ${\mathbb Z}_p$. Addition of two points (on an elliptic
curve over ${\mathbb Z}_n$), however, fails if and only if a factor of $n$ has
been found. The reason for this is that the procedure for adding points on
elliptic curves gives elements in ${\mathbb Z}_n$ and calculates the inverse
elements for these (with respect to multiplication mod $n$) in ${\mathbb Z}_n$.
The extended \index{Euclidean algorithm} Euclidean algorithm is used here. If
the addition of two points (that lie of an elliptic curve over ${\mathbb Z}_n$)
gives an element in ${\mathbb Z}_n$ that does not have an inverse element in
${\mathbb Z}_n$, then the extended Euclidean algorithm delivers a genuine factor
of $n$.

@@

>>>>

\newpage
\section{Elliptic Curves}
\index{Elliptic curves}
\hypertarget{ellcurve}{}
(Filipovics B. / B\"uger M. / Esslinger B. / Oyono R., April 2000, Updates: Dec. 2001, June 2002, Mar. 2003)

@@

There are factorisation algorithms based on elliptic curves
\footnote{In 1987 H.W. Lenstra published
a factorisation algorithm, based on elliptic curves (see \cite{Lenstra1987}).
The biggest compound number currently factorised\index{factoring records} with elliptic curves is 
the number $ 628^{59}-1, $ which has 55 decimal digits. It was
found Oct. 6th, 2001 by M. Izumi 
(See \hyperlink{Lenstra2}{ECMNET}\index{ECMNET}).
}. 
More precisely, these procedures exploit the fact that elliptic curves can
be defined over ${\mathbb Z}_n$ ($n$ composite number). Elliptic curves 
over ${\mathbb Z}_n$ do not form a group, because not every point on such 
an elliptic curve has an inverse point. This is connected with the fact 
that - if $n$ is a composite
number - there exist elements in ${\mathbb Z}_n$ that do not have an inverse
with respect to multiplication mod $n$. In order to add two points on an
elliptic curve over ${\mathbb Z}_n$, we can calculate in the same way as on
elliptic curves over ${\mathbb Z}_p$. Addition of two points (on an elliptic
curve over ${\mathbb Z}_n$), however, fails if and only if a factor of $n$ has
been found. The reason for this is that the procedure for adding points on
elliptic curves gives elements in ${\mathbb Z}_n$ and calculates the inverse
elements for these (with respect to multiplication mod $n$) in ${\mathbb Z}_n$.
The extended \index{Euclidean algorithm} Euclidean algorithm is used here. If
the addition of two points (that lie of an elliptic curve over ${\mathbb Z}_n$)
gives an element in ${\mathbb Z}_n$ that does not have an inverse element in
${\mathbb Z}_n$, then the extended Euclidean algorithm delivers a genuine factor
of $n$.

@@

   \item Key length comparison by Arjen Lenstra and Eric Verheul
         \href{http://cryptosavvy.com/table.htm}
	  {\texttt{http://cryptosavvy.com/table.htm}}

### 2003-03-27 script/en/chapters/introduction.tex
more be updates

Starting in the year 2000 this script became a part of the 
CrypTool\index{CrypTool} package. It should accomplish the program 
CrypTool by explaining some mathematical topics in more details, 
but still in a way which is easy to.

In order to also enable independent developers/authors work together 
the topics have been split up and for each topic an extra chapter was 
written readable by its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes, where you

can find the according 
functions within the CrypTool\index{CrypTool} program. Naturally 
there are much more interesting topics in mathematics and cryptography 
which could be more expanded -- therefore this is only one of many
ways to do it.

@@

This edition of the script updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes), 
  \item progress in number theory ("Primes in P"), 
  \item the factorisation of big numbers (TWIRL) and 
  \item progress in cryptanalysis of the AES standard. 
% \item usage of elliptic curves [not only over GP(p)].
\end{itemize}

@@

I'd be more than happy if this keeps on going also in the further 
open source versions of CrypTool\index{CrypTool}.

@@

This script is delivered together with CrypTool\index{CrypTool}. \par \vskip + 3pt

>>>>

Starting in the year 2000 this script became a part of the 
CrypTool\index{CrypTool} package. It should accomplish the program 
CrypTool by explaining some mathematical topics in more details, 
but still in a way which is easy to understand.

In order to also enable developers/authors to work independently together 
the topics have been split up and for each topic an extra chapter was 
written readable on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes, where you
can find the according 
functions within the CrypTool\index{CrypTool} program. Naturally 
there are much more interesting topics in mathematics and cryptography 
which could be more expanded -- therefore this is only one of many
ways to do it.

@@

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ (chap. 2.6), 
  \item progress in number theory ("Primes in P") (chap. 3.11.3), 
  \item the factorisation of big numbers (TWIRL) (chap. 3.11.3),
  \item progress in cryptanalysis of the AES standard (chap. 1.1) and
  \item progress in brute-force attacks against symmetric algorithms (chap. 1.1).
\end{itemize}

Chapter 6 of this script was also enhanced: now we explain in more details how elliptic curves are defined over different number fields.

@@

I'd be more than happy if this keeps on going also in the further 
open source versions of CrypTool\index{CrypTool} (they will be delivered by the University of Darmstadt).

@@

This script is delivered together with the program CrypTool\index{CrypTool}. \par \vskip + 3pt
